---
title: "Pericytes"
author: "Miguel Blanco"
format: html
editor: visual
---

### Load Libraries

```{r}
suppressPackageStartupMessages({
library(SeuratData)
library(tidyverse)
library(Seurat)
library(ggplot2)
library(ggrepel)
library(tibble)
library(DESeq2)
library(patchwork)
library(clustree)
library(fgsea)
})
```

```{r}
base_dir <- "/home/projects/dtu_00062/people/migchi/Pericytes/HealthyVs/"

pericytes <- readRDS(file = file.path(base_dir, "Pericytes_healthyVs.rds"))


```

### 1. Run Seurat Workflow (PCA, UMAP, Integration and Clustering)

```{r}
pericytes <- JoinLayers(pericytes)
```

```{r}
pericytes[["RNA"]] <- split(pericytes[["RNA"]], f = pericytes$Sequencing_Batch)
```

### 1.1 Rerun PCA and Integration

```{r}
pericytes <- NormalizeData(pericytes, normalization.method = "LogNormalize", scale.factor = 10000, verbose = FALSE, layer = "counts*")
pericytes <- FindVariableFeatures(pericytes, selection.method = "vst", layer = "data*")
pericytes <- ScaleData(pericytes, verbose = F)
pericytes <- RunPCA(pericytes)
```

```{r}
dim(Embeddings(pericytes, reduction = "pca"))
```

```{r}
PCApercentVarianceExplained <- function(Seurat_object)
  {
    std_devpc <- Seurat_object@reductions$pca@stdev
    pc_explainedvariance <- (std_devpc^2) / sum(std_devpc^2) * 100
    print(cumsum(pc_explainedvariance))
}

PCApercentVarianceExplained(pericytes)

```

```{r}
dim(Embeddings(pericytes, reduction = "pca"))

pc_scores <- Embeddings(pericytes, "pca")[, 1:2]   # first 2 PCs
pc_scores <- as.data.frame(pc_scores)
pc_scores$patient <- pericytes$Patient_studycase


patient_coords <- pc_scores |> 
  group_by(patient) |> 
  summarise(PC1 = mean(PC_1),
            PC2 = mean(PC_2)) |> 
  mutate(condition = case_when(grepl("Healthy", patient) ~ "Healthy",
                               TRUE ~ "Inflamed"))




PCAplot <- ggplot(patient_coords, aes(x = PC1, y = PC2, color = condition)) +
  geom_point(size = 4) +
  geom_text(aes(label = patient), vjust = -1, size = 3, show.legend = FALSE) +
  theme_classic() +
  labs(x = "PC1 (17.92%)", y = "PC2 (9.78%)",
       title = "Pericytes PCA") +
  scale_color_manual(
    values = c("Healthy" = "#1f77b4",   
               "Inflamed" = "#d62728")) +
  theme(plot.title = element_text(face = "bold",
                                  hjust = 0.5))



ggsave(PCAplot, file = "HealthyVs/PCAplot.png", width = 12, height = 10, dpi = 400)
```

### 1.1.2 Run MDS

```{r}
#Extract N components coordinates for all the cells
pca_coords <- Embeddings(pericytes, reduction = "pca")[, 1:10]


#Create a dataframe with that coordinates and also a "patient" column that indicates the patient to who each cell belongs
pca_df <- data.frame(pca_coords,
                     patient = pericytes$Patient,
                     Patient_Inflammation = pericytes$Patient_studycase)

#Calculate aggregate coordinates for each PC for each patient

centroids <- pca_df |> 
  #group_by(patient) |> 
  group_by(Patient_Inflammation) |> 
  summarise(across(starts_with("PC"), mean)) |> 
  ungroup()


# Compute Euclidean distances between patient centroids
dist_matrix <- dist(centroids |> dplyr::select(starts_with("PC")))

# Classical MDS
mds_coords <- cmdscale(dist_matrix, k = 2)

# Combine into data frame
mds_df <- data.frame(mds_coords, 
                    #Patient = centroids$patient,
                     Patient_Inflammation = centroids$Patient_Inflammation)

colnames(mds_df)[1:2] <- c("MDS1", "MDS2")


mds_df$Color <- case_when(
  grepl("^P.*inflamed_CD", mds_df$Patient_Inflammation) ~ "P_Inflamed",
  grepl("^P.*Healthy_CRC", mds_df$Patient_Inflammation) ~ "P_Healthy",
  grepl("^H.*Non_inflamed_CD", mds_df$Patient_Inflammation) ~ "H_NonInflamed",
  grepl("^H.*Healthy_CRC", mds_df$Patient_Inflammation) ~ "H_NonInflamed",
  TRUE ~ "H_Inflammed"
)


mds <- ggplot(mds_df, aes(x = MDS1, y = MDS2, label = Patient_Inflammation, color = Color)) +
  geom_point(size = 4) +
  geom_text(vjust = -0.8, size = 3) +
  theme_minimal() +
  labs(title = "MDS of Patient PCA Centroids",
       subtitle = "Based on average PC1–PC10 profiles")

ggsave(mds, filename = "mds.png", width = 12, height = 10, dpi = 400)
mds
```

#### 1.2 Run UMAPs and Integration

```{r}
pericytes <- FindNeighbors(pericytes, dims = 1:24, reduction = "pca")
pericytes <- FindClusters(pericytes, resolution = 0.5, cluster.name = "unintegrated_clusters")
```

```{r}
pericytes <- RunUMAP(pericytes, reduction = "pca", dims = 1:25, reduction.name = "unintegrated_UMAP")

DimPlot(pericytes, reduction = "unintegrated_UMAP", group.by = "Sequencing_Batch")
DimPlot(pericytes, reduction = "unintegrated_UMAP", group.by = "unintegrated_clusters")
```

#### 1.2.1 Integrate Layers

```{r, warning=FALSE}
pericytes <- IntegrateLayers(pericytes, method = HarmonyIntegration, 
                orig.reduction = "pca", 
                new.reduction = "harmony.integrated",
                group.by = "Sequencing_Batch",
                verbose = FALSE)
```

```{r}
pericytes <- FindNeighbors(pericytes, reduction = "harmony.integrated", dims = 1:25, verbose = FALSE)

```

```{r}
pericytes <- RunUMAP(pericytes, reduction = "harmony.integrated", dims = 1:25, reduction.name = "harmony_UMAP", verbose = F)

p1 <- DimPlot(pericytes, reduction = "unintegrated_UMAP",  group.by = "Sequencing_Batch")
p2 <- DimPlot(pericytes, reduction = "harmony_UMAP",  group.by = "Sequencing_Batch")

wrap_plots(p1,p2, ncol = 2, byrow = F)
p1
p2
```

```{r}
saveRDS(pericytes, file = "HealthyVs/Pericytes_healthyVs.rds")
```

```{r}
pericytes <- readRDS(file = "HealthyVs/Pericytes_healthyVs.rds")
```

### 1.3 Decide cluster resolution

```{r}
pericytes <- FindClusters(pericytes, resolution = 0.1, cluster.name = "harmony.clusters.0.1", verbose = FALSE )
pericytes <- FindClusters(pericytes, resolution = 0.15, cluster.name = "harmony.clusters.0.15", verbose = FALSE )
pericytes <- FindClusters(pericytes, resolution = 0.3, cluster.name = "harmony.clusters.0.3", verbose = FALSE )
pericytes <- FindClusters(pericytes, resolution = 0.5, cluster.name = "harmony.clusters.0.5", verbose = FALSE )
pericytes <- FindClusters(pericytes, resolution = 0.7, cluster.name = "harmony.clusters.0.7", verbose = FALSE )
```

```{r, warning=FALSE}


p0 <- DimPlot(pericytes, reduction = "harmony_UMAP",  group.by = "harmony.clusters.0.1", label = T)
p1 <- DimPlot(pericytes, reduction = "harmony_UMAP",  group.by = "harmony.clusters.0.15", label = T)
p2 <- DimPlot(pericytes, reduction = "harmony_UMAP",  group.by = "harmony.clusters.0.3", label = T)
p3 <- DimPlot(pericytes, reduction = "harmony_UMAP",  group.by = "harmony.clusters.0.5", label = T)

p <- wrap_plots(p0, p1, p2, p3, ncol = 2, nrow = 2, byrow = F)

ggsave("HealthyVs/pericytes_clust_resUMAP.png", plot = p, width = 15, height = 9, dpi = 400)
p
```

```{r}
clustree(pericytes, prefix = "harmony.clusters.")

```

Res 0.15 is the one that represents better our scenario. However, I think clusters 2 and 5 should be separated like in res 0.3. Let's change that manually

```{r}
pericytes@meta.data <- pericytes@meta.data |> 
  mutate(harmony.clusters.0.15 = case_when(harmony.clusters.0.3 == "5" ~ "6",
                                           TRUE ~ harmony.clusters.0.15))
```

```{r}
p1 <- DimPlot(pericytes, reduction = "harmony_UMAP",  group.by = "harmony.clusters.0.15", label = T)
p1
```

Cluster 3 and 8 together?

```{r, warning=FALSE}
p <- DimPlot(pericytes, reduction = "harmony_UMAP", 
             group.by = "harmony.clusters.0.15", 
             split.by = "study_case",
             label = T)

ggsave("HealthyVs/pericytes_splitUMAP.png", plot = p, width = 15, height = 9, dpi = 400)
p
```

```{r}
Idents(pericytes) <- "harmony.clusters.0.15"
```

Looks like the we have more cells in the lower part of the UMAP in inflamed samples (more populated clusters 1, 3 and 8) while in non inflamed samples we seem to have a higher relative abundance in cluster 4

```{r}
pericytes <- JoinLayers(pericytes)
```

## 2. Run DEG (Pseudobulk Approach)

### 2.1 Get aggregate expression per patient

```{r}
counts <- AggregateExpression(object = pericytes,
                              assays = "RNA",
                              group.by = "Patient_studycase",
                              return.seurat = F)
```

```{r}
counts <- counts$RNA
counts <- as.data.frame(counts)
```

```{r}
colData <- data.frame(patient = rownames(t(counts)))

colData <- colData |>
  mutate(study_case = case_when(
    grepl("Non-inflamed-CD", patient) ~ "Non_inflamed_CD",
    grepl("inflamed-CD", patient) ~ "inflamed_CD",
    grepl("Healthy-CRC", patient) ~ "Healthy_CRC"
  )) |>
  mutate(patient = sub("-.*", "", patient)) |>   # extract HH01 from HH01-inflamed-CD
  mutate(patient_studycase = paste0(patient, "_", study_case))      # get first letter (H or P)

rownames(colData) <- colnames(counts)
colData$batch <- pericytes@meta.data[rownames(colData), "Sequencing_batch"]
```

```{r}
library(dplyr)
colData <- tibble(sample = colnames(counts)) %>%
  mutate(
    study_case = case_when(
      grepl("Non-inflamed-CD", sample) ~ "Non_inflamed_CD",
      grepl("inflamed-CD",     sample) ~ "inflamed_CD",
      grepl("Healthy-CRC",     sample) ~ "Healthy_CRC",
      TRUE ~ NA_character_
    ), 
    patient = sub("-.*", "", sample),                 # "HH01" from "HH01-inflamed-CD"
  )
rownames(colData) <- colData$sample

colData <- colData |> 
  mutate(sample_key = gsub("-", "_", sample))

meta <- pericytes@meta.data %>%
  transmute(
    # build the same key you used in colData
    sample_key = paste0(Patient, "_", study_case),   # <-- adjust to your column names
    batch = Sequencing_Batch
  ) %>%
  filter(!is.na(sample_key), !is.na(batch)) %>%
  group_by(sample_key) %>%
  summarise(batch = names(sort(table(batch), decreasing = TRUE))[1], .groups = "drop") 
  # (mode of batch per sample_key; use first() if they're constant)


colData <- colData %>%
  left_join(meta, by = "sample_key")   # adds `batch`

# sanity checks
table(colData$study_case, colData$batch, useNA = "ifany")
sum(is.na(colData$batch))
```

```{r}
table(colData$study_case, colData$patient)
```

```{r}
colData$patient <- factor(colData$patient)
colData$study_case <- factor(colData$study_case)
colData$batch <- factor(colData$batch)
```

```{r}
dds <- DESeqDataSetFromMatrix(countData = counts, 
                              colData = colData, 
                              design = ~ study_case)

genes_to_keep <- rowSums(counts(dds)) >= 10
dds <- dds[genes_to_keep, ]
```

```{r}
table(colData$patient, colData$study_case)
```

```{r}
dds$study_case <- relevel(dds$study_case, ref = "Healthy_CRC")
dds <- DESeq(dds)
```

With this reference for the comparison, recall that log2FC \<0 = upregulated in inflammed

```{r}
resultsNames(dds)
res_HealthyVsInf <- results(dds, name = "study_case_inflamed_CD_vs_Healthy_CRC", alpha = 0.05 )
res_NIVshealthy <- results(dds, name = "study_case_Non_inflamed_CD_vs_Healthy_CRC", alpha = 0.05)
res_HealthyVsInf_df <- as.data.frame(res_HealthyVsInf)
res_NIVshealthy_df <- as.data.frame(res_NIVshealthy)
summary(res_HealthyVsInf)
summary(res_NIVshealthy)
```

```{r}
res_HealthyVsInf_df <- res_HealthyVsInf_df |> 
  rownames_to_column(var = "gene") |> 
  as_tibble() |> 
  arrange(padj)

res_NIVshealthy_df <- res_NIVshealthy_df |> 
  rownames_to_column(var = "gene") |> 
  as_tibble() |> 
  arrange(padj)
```

```{r}
sig_res_NIVshealthy_ <- res_NIVshealthy_df |> 
  filter(padj < 0.05) |> 
  arrange(desc(abs(log2FoldChange)))

sig_res_healthyVsInf <- res_HealthyVsInf_df |> 
  filter(padj < 0.05) |> 
  #arrange(padj)  
  arrange(desc(abs(log2FoldChange)))

dim(sig_res_NIVshealthy)
dim(sig_res_healthyVsInf)
head(sig_res_NIVshealthy)
head(sig_res_healthyVsInf)
```

Save tables

```{r}
write_csv(sig_res_NIVshealthy, file = "HealthyVs//sig_res_NIVshealthy.csv")
write_csv(sig_res_healthyVsInf, file = "HealthyVs//sig_res_healthyVsInf.csv")
```

### 2.2 Look for specific genes

```{r}
genes_specific <- c("VCAM1", "ICAM1", "PDGFB", "IL10", "CXCL1", "CCL2", "IL33", "CCL19", "CCL21", "TGFB1")

res_HealthyVsInf_df |> filter(gene %in% genes_specific)

sig_res_healthyVsInf |> filter(gene %in% genes_specific)
```

```{r}
VlnPlot(pericytes, features = c("VCAM1"), group.by = "study_case")

VlnPlot(pericytes, features = c("ICAM1"), group.by = "study_case") 
```

x

```{r}
sig_res_healthyVsInf |> filter(gene == "IL33")
```

### Volcano Plots

```{r}
all_genes = sig_res_healthyVsInf$gene
top_genes_healthy_INF <- head(sig_res_healthyVsInf, 20)
top_genes_healthy_padj <- head(sig_res_healthyVsInf |> arrange(padj), 5)

top_genes_healthy_INF <- bind_rows(top_genes_healthy_INF, top_genes_healthy_padj) |> distinct(gene, .keep_all = T)

relevant_genes <- c("IL7", "IL33", "IL17B", "MIF", "JAK3","CCL11", "CCL21", "CXCL2", "VCAM1", "CD44", "SERPING1", "CASP4", "PTGER1", "PTGER3", "PTGES", "PLA2G2A", "PLA2G5", "IGFBP4")

relevant_genes_df <- sig_res_healthyVsInf |> filter(gene %in% relevant_genes)

label_genes_df <- bind_rows(top_genes_healthy_INF, relevant_genes_df) %>%
  distinct(gene, .keep_all = TRUE)




Volcano_HealthyvsInf <- ggplot(sig_res_healthyVsInf, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = padj < 0.05 & abs(log2FoldChange) > 1), alpha = 0.6) +
  scale_color_manual(values = c("FALSE" = "blue", "TRUE" = "red"),
                     name = "Significant") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "grey") +
  geom_text_repel(data = label_genes_df,
                  aes(label = gene),
                  size = 3.5, max.overlaps = Inf,
                  segment.color = "black",
                  segment.size = 0.3,
                  arrow = arrow(length = unit(0.2, "npc"))) +
  labs(title = "Volcano Plot Healthy Vs Inflamed", x = "Log2 Fold Change", y = "-log10 adjusted p-value") +
  theme_minimal()+
  theme(plot.title = element_text(face = "bold"), hjust = 0.5)

Volcano_HealthyvsInf
```

```{r}

plot_df <- res_HealthyVsInf_df |>
  mutate(y = -log10(padj),
         color = case_when(padj < 0.05 & gene %in% relevant_genes ~ "TRUE & Inflammation Related",
                           padj < 0.05 ~ "TRUE",
                           TRUE ~ "FALSE")) |> 
  filter(is.finite(log2FoldChange), is.finite(y), padj > 0)

Volcano_HealthyINF_reshape <- ggplot(plot_df |> filter(!gene %in% c("SLN", "TUSC3", "PRRX1")), aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = color), alpha = 0.6) +
  scale_color_manual(values = c("FALSE" = "blue", "TRUE" = "red", "TRUE & Inflammation Related" = "#f1c40f"),
                     name = "Significant") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "grey") +
  geom_text_repel(data = label_genes_df |> filter(!gene %in% c("SLN", "TUSC3", "PRRX1")),
                  aes(label = gene),
                  size = 3.5, max.overlaps = Inf,
                  segment.color = "black",
                  segment.size = 0.3,
                  arrow = arrow(length = unit(0.2, "npc"))) +
  labs(title = "Volcano Plot Healthy Vs Inflamed", x = "Log2 Fold Change", y = "-log10 adjusted p-value") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))+
  coord_cartesian(xlim = c(-8, 8), ylim = c(0, 10.5))

Volcano_HealthyINF_reshape
```

```{r}
Volcano_HealthyINF_reshape_onlyrelevant <- ggplot(plot_df |> filter(!gene %in% c("SLN", "TUSC3", "PRRX1")), aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = color), alpha = 0.6) +
  scale_color_manual(values = c("FALSE" = "blue", "TRUE" = "red", "TRUE & Inflammation Related" = "#f1c40f"),
                     name = "Significant") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "grey") +
  geom_text_repel(data = relevant_genes_df,
                  aes(label = gene),
                  size = 3.5, max.overlaps = Inf,
                  segment.color = "black",
                  segment.size = 0.3,
                  arrow = arrow(length = unit(0.2, "npc"))) +
  labs(title = "Volcano Plot Healthy Vs Inflamed", x = "Log2 Fold Change", y = "-log10 adjusted p-value") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))+
  coord_cartesian(xlim = c(-8, 8), ylim = c(0, 10.5))

Volcano_HealthyINF_reshape_onlyrelevant
```

```{r}
#ggsave(filename = "HealthyVs/Volcano/volcano_NIvsHealthy.png", plot = Volcano_NIvsHealthy, width = 14, height = 12, dpi = 300)

ggsave(filename = "HealthyVs/Volcano/volcano_HealthyvsINF.png", plot = Volcano_HealthyvsInf, width = 14, height = 12, dpi = 300, bg = "white")

ggsave(filename = "HealthyVs/Volcano/volcano_HealthyvsINF_reshape.png", plot = Volcano_HealthyINF_reshape, width = 14, height = 12, dpi = 300, bg = "white")

ggsave(filename = "HealthyVs/Volcano/volcano_HealthyvsINF_reshape_onlyrelevant.png", plot = Volcano_HealthyINF_reshape_onlyrelevant, width = 14, height = 12, dpi = 300, bg = "white")
```

### HeatMap

```{r}
library(pheatmap)
# Extract gene names
heatmap_genes_CDINF <- head(sig_res_healthyVsInf$gene, 25)

# Extract VST-transformed counts for those genes
norm_counts_CDINF <- assay(vst(dds, blind = TRUE))


ann_col <- as.data.frame(colData(dds)[, "study_case", drop = FALSE])
rownames(ann_col) <- colnames(dds)

order <- order(ann_col$study_case)
norm_counts_CDINF <- norm_counts_CDINF[, order]

ann_col <- ann_col |> 
  arrange(study_case)
# Plot heatmap
heatmap_top25 <- pheatmap(norm_counts_CDINF, 
         cluster_rows = FALSE, 
         show_rownames = TRUE,
         cluster_cols = FALSE, 
         annotation_col = ann_col)

heatmap_top25
```

```{r}
library(pheatmap)
# Extract gene names
heatmap_genes_CDINF <- relevant_genes

# Extract VST-transformed counts for those genes
norm_counts_CDINF <- assay(vst(dds, blind = TRUE))[heatmap_genes_CDINF, ]


ann_col <- as.data.frame(colData(dds)[, "study_case", drop = FALSE])
rownames(ann_col) <- colnames(dds)

order <- order(ann_col$study_case)
norm_counts_CDINF <- norm_counts_CDINF[, order]

ann_col <- ann_col |> 
  arrange(study_case)
# Plot heatmap
heatmap_top25 <- pheatmap(norm_counts_CDINF, 
         cluster_rows = FALSE, 
         show_rownames = TRUE,
         cluster_cols = FALSE, 
         annotation_col = ann_col)

heatmap_top25






## PDGFR, VCAM and ICAM heatmap





heatmap_genes_PDGFR <- c("VCAM1", "ICAM1", "PDGFB")

heatmap_genes_PDGFR <- c("MIF", "CXCL1", "CCL2")


# Extract VST-transformed counts
norm_counts_PDGFR <- assay(vst(dds, blind = TRUE))[heatmap_genes_PDGFR, ]

# Create annotation dataframe
ann_col <- as.data.frame(colData(dds)[, "study_case", drop = FALSE])
rownames(ann_col) <- colnames(dds)

# Filter to specific patients
keep_samples <- rownames(ann_col)[ann_col$patient_studycase %in% Patients_studycase_tokeep]
inflamed_patients <- grep("inflamed", Patients_studycase_tokeep, value = TRUE)




# Subset both counts and annotation
norm_counts_PDGFR <- norm_counts_PDGFR[, keep_samples]
ann_col <- ann_col[keep_samples, , drop = FALSE]

ann_col$inflamed <- ifelse(ann_col$patient_studycase %in% inflamed_patients, "Inflamed", "Healthy")
ann_colors <- list(Highlight = c(Inflamed = "red", Healthy = "green"))

# Order by patient
order <- order(ann_col$study_case)
norm_counts_PDGFR <- norm_counts_PDGFR[, order]
ann_col <- ann_col[order, , drop = FALSE]

# Plot heatmap
heatmap_PDGFR <- pheatmap(norm_counts_PDGFR, 
                          cluster_rows = FALSE, 
                          show_rownames = TRUE,
                          cluster_cols = FALSE,
                          annotation_col = ann_col,
                          annotation_colors = ann_colors)
heatmap_PDGFR







```

```{r}
plot_deg_heatmap <- function(
  dds,
  sig_genes,
  sample_key_col = "sample_key",
  condition_col  = "study_case",
  collapse_to_patient = TRUE,
  use_vst = TRUE,
  subset_conditions = NULL,          
  split_by_condition = TRUE,         
  condition_levels = NULL,           
  palette = c("purpleYellow","blueWhiteRed"),
  z_limits = c(-2.5, 2.5),
  cond_colors = NULL,
  palette_seed = NULL,
  show_gene_names = F,
  row_name_fontsize = 7,
  file_pdf = NULL,                   # e.g. "DEG_heatmap_by_patient.pdf"
  file_png = NULL,                   # e.g. "DEG_heatmap_by_patient.png"
  png_width = 2000, png_height = 1600, png_res = 200
){
  stopifnot(inherits(dds, "DESeqDataSet"))
  palette <- match.arg(palette)

  suppressPackageStartupMessages({
    library(ComplexHeatmap); library(circlize); library(dplyr)
    library(DESeq2)
  })

  ## --- matrix log-like ---
  mat <- if (use_vst) assay(vst(dds, blind = TRUE)) else {
    stop("Provide a log-like matrix yourself if use_vst = FALSE.")
  }
  
  
  cluster_rows_flag <- !isTRUE(show_gene_names) 

  ## --- Make sure only to include present genes ---
  genes_present <- intersect(sig_genes, rownames(mat))
  if (length(genes_present) < 5L)
    stop("Fewer than 5 significant genes found in expression matrix rownames.")
  mat <- mat[genes_present, , drop = FALSE]

  ## --- colData ---
  cd <- as.data.frame(colData(dds))
  req <- c(sample_key_col, condition_col)
  if (!all(req %in% names(cd)))
    stop("colData(dds) must contain: ", paste(req, collapse=", "))

  cd[[sample_key_col]] <- droplevels(as.factor(cd[[sample_key_col]]))
  cd[[condition_col]]  <- droplevels(as.factor(cd[[condition_col]]))

  ## Align columns
  cd <- cd[colnames(mat), , drop = FALSE]

  ## --- Colapse as avg per patient ---
  if (collapse_to_patient) {
    grp <- cd[[sample_key_col]]
    mat <- t(rowsum(t(mat), group = grp)) / as.vector(table(grp))

    
    patient_to_cond <- tapply(cd[[condition_col]], grp, function(v){
      vv <- as.character(v); names(sort(table(vv), decreasing = TRUE))[1]
    })
    top_annot_df <- data.frame(
      study_case = factor(patient_to_cond[ colnames(mat) ])
    )
    rownames(top_annot_df) <- colnames(mat)
  } else {
    top_annot_df <- data.frame(
      sample_key = cd[[sample_key_col]],
      study_case = cd[[condition_col]]
    )
    rownames(top_annot_df) <- rownames(cd)
  }

  ## --- filter conditions ---
  if (!is.null(subset_conditions)) {
    keep <- as.character(top_annot_df$study_case) %in% subset_conditions
    mat <- mat[, keep, drop = FALSE]
    top_annot_df <- droplevels(top_annot_df[keep, , drop = FALSE])
    if (ncol(mat) < 2) stop("After filtering, fewer than 2 columns remain.")
  }

  ## --- order by condition ---
  if (is.null(condition_levels)) {
    condition_levels <- levels(droplevels(as.factor(top_annot_df$study_case)))
  }
  top_annot_df$study_case <- factor(top_annot_df$study_case, levels = condition_levels)

  
  
  ## --- Z-score per gene ---
  mat_z <- t(scale(t(mat)))
  keep_cols <- !is.na(top_annot_df$study_case)
  if (!all(keep_cols)) {
    mat      <- mat[, keep_cols, drop = FALSE]      
    mat_z    <- mat_z[, keep_cols, drop = FALSE]    
  top_annot_df <- droplevels(top_annot_df[keep_cols, , drop = FALSE])
  }
  mat_z <- mat_z[keep_rows, , drop = FALSE]

  ## --- palettes ---
  col_fun <- switch(palette,
    purpleYellow = circlize::colorRamp2(
      c(z_limits[1], 0, z_limits[2]),
      c("#5B2A86", "#FFFFFF", "#FDE725")
    ),
    blueWhiteRed = circlize::colorRamp2(
      c(z_limits[1], 0, z_limits[2]),
      c("#2C7BB6", "#F7F7F7", "#D7191C")
    )
  )

  # Color per condition
  
  cond_levels <- levels(top_annot_df$study_case)
  
  if (!is.null(cond_colors)) {
    # expect a *named* vector: names must be condition labels
    if (!all(cond_levels %in% names(cond_colors)))
      stop("cond_colors must be a named vector covering: ",
           paste(cond_levels, collapse = ", "))
    cond_cols <- cond_colors[cond_levels]   # keep order of levels
  } else if (!is.null(palette_seed)) {
    set.seed(palette_seed)                  # reproducible random if you want it
    cond_cols <- setNames(
      circlize::rand_color(length(cond_levels), luminosity = "bright"),
      cond_levels
    )
  } else {
    # deterministic default (Okabe–Ito, color-blind friendly)
    okabe <- c("#D55E00","#0072B2","#009E73","#CC79A7",
               "#F0E442","#56B4E9","#E69F00","#000000")
    if (length(cond_levels) > length(okabe))
      stop("Provide cond_colors for >8 groups.")
    cond_cols <- setNames(okabe[seq_along(cond_levels)], cond_levels)
  }

  ha <- HeatmapAnnotation(
  Condition = top_annot_df$study_case,
  col = list(Condition = cond_cols),
  annotation_height = grid::unit(5, "mm"),
  gp = grid::gpar(col = NA),
  annotation_name_gp = grid::gpar(fontface = "bold"),  # estilo del nombre
  annotation_legend_param = list(Condition = list(title = "Condition"))
)

  ## --- Order Columns ---
  column_order <- seq_len(ncol(mat_z))
  if (!split_by_condition) {
    # By condition and then by cluster
    hc_cols <- hclust(as.dist(1 - cor(mat_z, method = "pearson")), method = "average")
    glob_ord <- hc_cols$order
    idx_by_cond <- split(seq_len(ncol(mat_z)), top_annot_df$study_case)
    column_order <- unlist(lapply(levels(top_annot_df$study_case), function(cl){
      idx <- idx_by_cond[[cl]]; idx[order(match(idx, glob_ord))]
    }))
    mat_z        <- mat_z[, column_order, drop = FALSE]
    top_annot_df <- top_annot_df[column_order, , drop = FALSE]
    ha <- HeatmapAnnotation(
  Condition = top_annot_df$study_case,
  col = list(Condition = cond_cols),
  annotation_height = grid::unit(5, "mm"),
  gp = grid::gpar(col = NA),
  annotation_name_gp = grid::gpar(fontface = "bold"),  # estilo del nombre
  annotation_legend_param = list(Condition = list(title = "Condition"))
)
  }

  ## --- Build heatmap ---
  ht <- Heatmap(
    mat_z,
    name = "Z-score",
    col = col_fun,
    top_annotation = ha,
    show_row_names = show_gene_names,
    show_column_names = TRUE,
    column_names_gp = grid::gpar(fontsize = 8),

    cluster_rows = cluster_rows_flag,
    cluster_columns = split_by_condition,   # si no hay split, fijamos orden manual
    clustering_distance_rows = "pearson",
    clustering_distance_columns = "pearson",

    column_split = if (split_by_condition) top_annot_df$study_case else NULL,
    cluster_column_slices = if (split_by_condition) FALSE else TRUE,
    column_gap = if (split_by_condition) grid::unit(2.5, "mm") else grid::unit(0, "mm"),

    use_raster = TRUE,
    raster_quality = 2,
    heatmap_legend_param = list(at = c(z_limits[1], 0, z_limits[2]), title = "Z-score")
  )

  ## --- guardar si se pide ---
  if (!is.null(file_pdf)) {
    pdf(file_pdf, width = if (collapse_to_patient) 10 else 14, height = 10)
    draw(ht)
    dev.off()
  }
  if (!is.null(file_png)) {
    png(file_png, width = png_width, height = png_height, res = png_res)
    draw(ht)
    dev.off()
  }

  return(ht)
}
```

```{r}


ht1 <- plot_deg_heatmap(
  dds, sig_genes = sig_res_healthyVsInf$gene,
  collapse_to_patient = TRUE,
  split_by_condition = F,
  condition_levels = c("inflamed_CD","Non_inflamed_CD","Healthy_CRC"),
  palette = "blueWhiteRed",
  file_png = "HealthyVs/Heatmap/HT_Nonsplitted_allconditions.png"
)

ht1

ht2 <- plot_deg_heatmap(
  dds, sig_genes = sig_res_healthyVsInf$gene,
  collapse_to_patient = TRUE,
  split_by_condition = F,
  condition_levels = c("inflamed_CD","Healthy_CRC"),
  file_png = "HealthyVs/Heatmap/HT_Nonsplitted_HandInf.png",
  palette = "blueWhiteRed"
)

ht2

ht3 <- plot_deg_heatmap(
  dds, sig_genes = sig_res_healthyVsInf$gene,
  collapse_to_patient = TRUE,
  split_by_condition = T,
  condition_levels = c("inflamed_CD","Non_inflamed_CD", "Healthy_CRC"),
  palette = "blueWhiteRed",
  file_png = "HealthyVs/Heatmap/HT_splitted_allconditions.png"
)

ht3


ht4 <- plot_deg_heatmap(
  dds, sig_genes = sig_res_healthyVsInf$gene,
  collapse_to_patient = TRUE,
  subset_conditions = c("inflamed_CD", "Healthy_CRC"),
  split_by_condition = T,
  condition_levels = c("inflamed_CD","Healthy_CRC"),
  palette = "blueWhiteRed",
  file_png = "HealthyVs/Heatmap/HT_splitted_HandInf.png",
)

ht4


ht5 <- plot_deg_heatmap(
  dds, sig_genes = relevant_genes_df$gene,
  collapse_to_patient = TRUE,
  split_by_condition = F,
  condition_levels = c("inflamed_CD","Healthy_CRC"),
  palette = "blueWhiteRed",
  show_gene_names = T,
  file_png = "HealthyVs/Heatmap/HT_Nonsplitted_HandInf_relevant_genes.png",
)


ht5
```

```{r}
table(pericytes$Patient_studycase, pericytes$Sequencing_Batch)
```

## 3. FGSEA

```{r}
library(org.Hs.eg.db)
library(AnnotationDbi)
Entrez <- AnnotationDbi::select(org.Hs.eg.db,
                                keys = res_HealthyVsInf_df$gene,
                                columns = "ENTREZID",
                                keytype = "SYMBOL")

Entrez <- as_tibble(Entrez)

res_HealthyVsInf_df <- inner_join(res_HealthyVsInf_df, Entrez, by = c("gene" = "SYMBOL"))

res_HealthyVsInf_df <- res_HealthyVsInf_df |> 
  filter(!is.na(ENTREZID), is.finite(pvalue), is.finite(log2FoldChange), pvalue > 0) |> 
  mutate(ranked_score =  (-log10(pvalue) * sign(log2FoldChange)))

rankdata <-  res_HealthyVsInf_df$ranked_score 
names(rankdata) <- res_HealthyVsInf_df$ENTREZID



```

```{r}
kegg_msigdb <- msigdbr(species = "Homo sapiens", category = "C5", subcategory = "BP")

kegg_list <- kegg_msigdb |> 
  split(x = kegg_msigdb$entrez_gene, f = kegg_msigdb$gs_name)

rctome_pathways <- reactomePathways(names(rankdata))
```

```{r}


fgsea_NIvsINF <- fgsea(pathways = kegg_list, stats = rankdata)

fgsea_NIvsINF_reactome <- fgsea(pathways = rctome_pathways, stats = rankdata)



```

### 3.1 Collapse Pathways

```{r}
collapsedPathways <- collapsePathways(fgsea_NIvsINF[order(pval)][padj < 0.01], 
                                      kegg_list, rankdata)
mainPathways <- fgsea_NIvsINF[pathway %in% collapsedPathways$mainPathways][
                         order(-NES), pathway]

```

Extract top pathways

```{r}
view(fgsea_NIvsINF_reactome |> filter(padj < 0.05))
```

```{r}
extract_topbot_pathways <- function(fgsea_result, padj_threshold = 0.05, top_n = 10, bottom_n = 10, total_n = 20) {
  fgsea_tidy <- as_tibble(fgsea_result)
  
  top <- fgsea_tidy |>
    filter(padj < padj_threshold) |>
    arrange(desc(NES)) |>
    head(top_n)
  
  bottom <- fgsea_tidy |>
    filter(padj < padj_threshold) |>
    arrange(NES) |>
    head(bottom_n)
  
  topabs <- fgsea_tidy |> 
    filter(padj < padj_threshold) |> 
    arrange(desc(abs(NES))) |> 
    head(total_n)
  
  bot20 <- fgsea_tidy |>
    filter(padj < 0.1) |>
    arrange(NES) |>
    head(total_n)
  
  topbot <- bind_rows(top, bottom)
  return(list(topabs = topabs , topbot = topbot, bot20 = bot20, bottom = bottom))
}




Kegg_top <- extract_topbot_pathways(fgsea_NIvsINF)
topbot_KEGG <- Kegg_top$topbot
topabs_KEGG <- Kegg_top$topabs



Rctome_top <- extract_topbot_pathways(fgsea_NIvsINF_reactome |>  filter(pathway != "Sensory processing of sound"))
topbot_Rctome <- Rctome_top$topbot
topabs_Rctome <- Rctome_top$topabs
bot20_Rctome <- Rctome_top$bot20
bot10_Rctome <- Rctome_top$bottom


```

```{r}
interesting_pathways <- c("Interleukin-12 family signaling", "Collagen degradation", "Complement cascade", "Cytokine Signaling in Immune system", "Degradation of the extracellular matrix", "Interferon Signaling", "Modulation of host responses by IFN-stimulated genes", "Interferon gamma signaling", "Interleukin-4 and Interleukin-13 signaling", "Signaling by Interleukins", "Immune System", "Innate Immune System", "Antigen processing: Ub, ATP-independent proteasomal degradation", "Synthesis of Prostaglandins (PG) and Thromboxanes (TX)", "Modulation of host responses by IFN-stimulated genes")


interesting_pathways_all <- c("Interleukin-12 family signaling", "Collagen degradation", "Complement cascade", "Cytokine Signaling in Immune system", "Degradation of the extracellular matrix", "Interferon Signaling", "Modulation of host responses by IFN-stimulated genes", "Interferon gamma signaling", "Interleukin-4 and Interleukin-13 signaling", "Signaling by Interleukins", "Immune System", "Innate Immune System", "Antigen processing: Ub, ATP-independent proteasomal degradation", "Synthesis of Prostaglandins (PG) and Thromboxanes (TX)", "Modulation of host responses by IFN-stimulated genes", "Cell-Cell communication", "VEGFA-VEGFR2 Pathway")


Rctome_interesting <- fgsea_NIvsINF_reactome |>  filter(pathway %in% interesting_pathways)

Rctome_interesting_all <- fgsea_NIvsINF_reactome |>  filter(pathway %in% interesting_pathways_all)
```

### 3.2 FGSEA Barplot

```{r}
library(ggplot2)
library(scales)

plot_fgsea_bar <- function(data, plot_title, y_label) {
  ggplot(data, aes(x = NES, y = reorder(pathway, NES))) +
    geom_col(aes(fill = -log10(padj))) +
    scale_fill_gradientn(
      colors = c("#2166ac", "#67a9cf", "#d1e5f0", "#ef8a62", "#b2182b", "#b2182b"),
      values = rescale(c(1.3, 2, 5, 10, 13)),  # -log10(padj)
      limits = c(0.05, 15),
      breaks = c(1, 3, 5, 10, 15),
      labels = c("0.05", "0.01", "1e-5", "1e-10", "1e-13"),
      name = "padj"
    ) +
    labs(
      title = plot_title,
      x = "NES", y = y_label
    ) +
    theme_minimal() +
    theme(
      axis.title = element_text(size = 10, face = "bold"),
      axis.text.y  = element_text(size = 10, face = "bold"),
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
      legend.title = element_text(size = 9, face = "bold"),
      legend.text = element_text(size = 8)
    )
}



plot_fgsea_bar <- function(data, plot_title, y_label) {
  library(ggplot2); library(scales)

  # 1) Clamp zeros to a tiny positive value to avoid Inf
  eps <- 1e-300
  data$padj_clamped <- pmax(data$padj, eps)

  # 2) Work on log scale
  data$log_padj <- -log10(data$padj_clamped)

  # Integer breaks on the log scale
  rng <- range(data$log_padj[is.finite(data$log_padj)], na.rm = TRUE)
  breaks <- seq(floor(rng[1]), ceiling(rng[2]), by = 1)

  # Positions for gradientn must be in [0,1]
  values <- rescale(breaks, to = c(0, 1), from = range(breaks))

  ggplot(data, aes(x = NES, y = reorder(pathway, NES))) +
    geom_col(aes(fill = log_padj)) +
    scale_fill_gradientn(
      colors = c("#2166ac", "#67a9cf", "#d1e5f0", "#ef8a62", "#b2182b", "#67001f"),
      values = values,
      breaks = breaks,
      labels = label_scientific()(10^(-breaks)),  # 1e-2, 1e-3, ...
      name   = "padj"
    ) +
    labs(title = plot_title, x = "NES", y = y_label) +
    theme_minimal(base_size = 10) +
    theme(plot.title = element_text(face = "bold", hjust = 0.5),
          axis.text.y  = element_text(size = 10, face = "bold", family = "Arial"),
          legend.title = element_text(face = "bold"),
          plot.title.position = "plot")
}
```

```{r}




fgsea_plot_topbot10 <- plot_fgsea_bar(topbot_KEGG, plot_title = "Top 10 Most Up/Downregulated KEGG Pathways in Inflamed vs Healthy", y_label = "Pathway (KEGG)")
fgsea_plot_topbot10
fgsea_plot_top20 <- plot_fgsea_bar(topabs_KEGG, plot_title = "Top 20 most disregulated KEGG pathways in Inflamed vs Healthy (log2FC)", y_label = "Pathway (KEGG)")
fgsea_plot_top20



fgsea_plot_topbot10_Rctome <- plot_fgsea_bar(topbot_Rctome, plot_title = "Top 10 Most Up/Downregulated Reactome Pathways in Inflamed vs Healthy", y_label = "Pathway (Reactome)")
fgsea_plot_topbot10_Rctome
fgsea_plot_top20_Rctome <- plot_fgsea_bar(topabs_Rctome, plot_title = "Top 20 most disregulated Reactome pathways in Inflamed vs Healthy", y_label = "Pathway (Reactome)")
fgsea_plot_top20_Rctome
fgsea_plot_bot20_Rctome <- plot_fgsea_bar(bot20_Rctome, plot_title = "Top 20 most Downregulated Reactome pathways in Inflamed vs Healthy", y_label = "Pathway (Reactome)")
fgsea_plot_bot20_Rctome


fgsea_plot_bot10_Rctome <- plot_fgsea_bar(bot10_Rctome, plot_title = "Top 10 most Downregulated Reactome pathways in Inflamed vs Healthy", y_label = "Pathway (Reactome)")
fgsea_plot_bot10_Rctome


fgsea_interesting_pathways <- plot_fgsea_bar(Rctome_interesting, plot_title = "IBD-related Up-regulated Reactome Pathways in Inflamed vs Healthy", y_label = "Pathway (Reactome)")
fgsea_interesting_pathways

fgsea_interesting_pathways_all <- plot_fgsea_bar(Rctome_interesting_all, plot_title = "IBD-related Reactome Pathways in Inflamed vs Healthy", y_label = "Pathway (Reactome)")
fgsea_interesting_pathways_all
```

```{r, warning=FALSE}



ggsave(plot = fgsea_plot_top20, filename = file.path(base_dir, "fgsea/fgsea_barplot_top20abs_KEGG.png"), width = 14, height = 10, dpi = 400, bg = "white")
ggsave(plot = fgsea_plot_topbot10, filename = file.path(base_dir, "fgsea/fgsea_barplot_topbot10_KEGG.png"), width = 14, height = 10, dpi = 400, bg = "white" )


ggsave(plot = fgsea_plot_top20_Rctome, filename = file.path(base_dir, "fgsea/fgsea_barplot_top20abs_rctome.png"), width = 14, height = 10, dpi = 400, bg = "white" )
ggsave(plot = fgsea_plot_topbot10_Rctome, filename = file.path(base_dir, "fgsea/fgsea_barplot_topbot10_rctome.png"), width = 14, height = 10, dpi = 400, bg = "white" )
ggsave(plot = fgsea_plot_bot20_Rctome, filename = file.path(base_dir, "fgsea/fgsea_barplot_bot20abs_rctome.png"), width = 14, height = 10, dpi = 400, bg = "white" )

ggsave(plot = fgsea_plot_bot10_Rctome, filename = file.path(base_dir, "fgsea/fgsea_barplot_bot10abs_rctome.png"), width = 14, height = 10, dpi = 400, bg = "white" )


ggsave(plot = fgsea_interesting_pathways, filename = file.path(base_dir, "fgsea/fgsea_barplot_IBD_rctome_reshaped.png"), width = 14, height = 10, dpi = 400, bg = "white" )


ggsave(plot = fgsea_interesting_pathways_all, filename = file.path(base_dir, "fgsea/fgsea_barplot_IBD_rctome_reshaped_all.png"), width = 14, height = 10, dpi = 400, bg = "white" )


```

Almost all the interesting (Potentially related to IBD/Inflammation) pathways spotted in the analysis are downregulated in Inflamed samples. Some of them are:

**ERK/MAPK targets & MAPK pathways:** Wound healing, angiogenesis, cross-talk with immune cells

**VEGFA/VEGFR2:** May suggest vascular inestability, which added to the overexpression of CD44 and VCAM-1 would suggest an enhanced migration of immune cells from the blood to the tissue, in which pericytes would facilitate the process by dismantling the endothelium junctions and secrete adhesion molecules (like CD44 and VCAM-1).

**Cell-Cell communication**: Important, can potentially reduce anti-inflammatory signaling pathways to inmune cells, fibroblasts... Will be further explored

**Platelet Homeostasis and sensitization:** Lower platelet adhesion can lead to coagulation imbalances or vascular leakiness (IBD features)

**Inositol phosphate metabolism:** Intracelular calcium signaling and cell motility, which can affect pericyte contraction and vascular tone

**FGFR pathways**: INvolved in angiogenesis, wound repair and epithelial homeostasis

**PI3K**: Mediate survival signal and anti-inflammatory responses

## 
